---
title: "tsg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tsg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tsg)
```

The `tsg` package provides a set of functions for generating frequency tables, cross-tabulations, and formatted output tables. It is designed to work with data frames and supports various customization options.

## Generate frequency table 

The `generate_frequency()` function creates frequency tables for one or more categorical variables in a data frame. It supports a variety of enhancements, such as sorting, adding totals and percentages, handling missing values, and customizing labels. This function is highly versatile and can work with grouped data, outputting either a single table or a list of tables.

### Basic usage

### 


## Generate cross-tabulation 

Using the `generate_crosstab()` function

This function creates a cross-tabulation (contingency table) from the data, allowing for analysis of the relationship between two or more categorical variables.


## Generate output using the `generate_output()` function

This function generates a formatted output table from the frequency or cross-tabulation data. It supports various formatting options and can include metadata such as titles and subtitles.
